Open-source landscape for 3D CSEM modelling

### 2020-04-08 manuscript/Manuscript.tex
Clean-up intro

Controlled-source electromagnetic (CSEM) measurements is a frequently applied
method in exploration geophysics for various applications, such as ground
water, geothermal, oil and gas, mining, civil engineering, or geo-hazards.
Modelling these electromagnetic fields is therefore of great interest in order
to understand the measured data. Publications regarding 3D modelling in
electromagnetic methods started to appear as early as the 1970's and 1980's.
These early publications where all integral equation (IE) methods, having an
anomaly embedded within a layered medium, for loop-loop type transient EM
measurements \citep{GJI.74.Raiche, GEO.75.Hohmann, GJI.82.Das, GEO.86.Newman}
and magnetotelluric (MT) measurements \citep{GEO.84.Wannamaker}.

In the 1990's computer became sufficiently powerful that 3D modelling gained
traction, which resulted amongst other in the publication of the book
\emph{Three-Dimensional Electromagnetics} \citep{B.SEG.99.Oristaglio} by the
SEG. Often cited publications from that time are \cite{RSC.94.Mackie}, 3D MT
computation; \cite{RS.94.Druskin}, frequency- and time-domain modelling using a
Yee grid and a global Krylov subspace approximation; and \cite{RS.96.Alumbaugh,
GJI.97.Newman}, low-to-high frequency computation on massively parallel
computers.

The continuous improvement of computing power and the CSEM boom in the early
2000's in the hydrocarbon industry led to a wealth of publications. The amount
of available numerical solutions can be overwhelming and is part of the reason
why there are hundreds of publications about the topic. There are the different
methods to solve Maxwell's equation, such as the IE method
\citep{GJI.74.Raiche, RS.02.Hursan, GEO.06.Zhdanov, GP.10.Tehrani,
CAG.16.Kruglyakov, MGS.17.Kruglyakov} and different variations of the
differential equation (DE) method, for instance finite differences (FD)
\citep{IEEE.66.Yee, RSC.94.Mackie, RS.94.Druskin, GEO.09.Streich,
CAG.13.Sommer}, finite elementes (FE) \citep{GJI.11.Schwarzbach,
GEO.12.daSilva, GJI.13.Puzyrev, GJI.13.Grayver, SEG.16.Zhang}, finite volume
(FV) \citep{EM.90.Madsen, ECP.07.Haber, GEO.14.Jahandari}, and finite
integration technique (FIT) \citep{PIER.01.Clemens, GP.06.Mulder}. And these
are just the most common ones.

@@

A very well written overview up to the year 2005 of the different approaches to
3D EM modelling is given by \cite{SG.05.Avdeev}. In the last 15 years the
publications with regards to 3D EM  modelling grew tremendously, driven
probably by improved computer powers. Avdeev finished his review with the
following statement: «\emph{The most important challenge that faces the EM
community today is to convince software developers to put their 3-D EM forward
and inverse solutions into the public domain, at least after some time. This
would have a strong impact on the whole subject and the developers would
benefit from feedback regarding the real needs of the end-users.}»

It is this topic that we want to address, open-source 3D CSEM codes. Some big
exploration companies and service companies have their own codes, as well as
certain research consortia. But how about codes that are open-source?
Traditionally research code was often available upon request by the author.
Codes distributed that way often come with the request to not share the code,
or they often come without a license attached at all (which means that they are
not open-source). Also, at times there can be a significant hurdle to install
a code.

However, the meaning of open-source evolved rapidly in the last decades. Today
open-source generally not only means that a code is available (with a proper
license). It includes much more today, such as that the code is hosted online,
is under version control, has the possibility to file issues and make pull
requests to fix bugs. Good codes also have continuous integration that includes
extensive testing of the code base and online hosted documentation that
includes the API. As such the term has shifted from purely open-source code
to development in the open and increasingly building a global community. There
are a number of projects within the realm of geophysical exploration, besides
the ones we present here, e.g. pyGIMLi \citep{CAG.17.Rucker} and Fatiando
\citep{JOSS.18.Uieda}.

Newer \todo{wrong term, think about it}programming environments  have also
brought much simplifications in the installation process and have most
importantly simplified it for code developers to make their codes easily
available and installable. All our presented codes are in the Python
environment, and each code can be install with a single command (which includes
downloading and installation).

~\\ \hrule

It is worth mentioning that there are many more open-source 3D electromagnetic
codes developed in other fields than geophysics. Examples include \dots (TODO:
add at least three references). While these codes could potentially be used for
the same goals as presented here we restrict our review to codes purpose-built
for geophysical applications.

An integral part of this publication are the necessary instructions and codes
to reproduce all published results. In fact, the provided codes contain more
than just the results shown here, as including everything here would have made
the article to lengthy.

This will significantly increase the user base and the number of people who are
actually doing 3D EM modelling.

~\\ \hrule

\begin{itemize}
  \item Why need:
    \begin{itemize}
      \item analytical solutions
      \item semi-analytical solutions such as 1D codes (analytical [recursion]
        in k-f domain followed by numerical transformation).
      \item need of comparison of more complex models.
    \end{itemize}
  \item What exists:
    \begin{itemize}
      \item very quick run-down of codes, methods, discretizations.
      \item Thesis Raphael
      \item LitRev Dieter
    \end{itemize}
  \item What changed:
    \begin{itemize}
      \item Python
      \item Demand/Desire for reproducible research
      \item Open-source has changed:
        \begin{itemize}
          \item easy install
          \item documented
          \item bug/issues
          \item communities
          \item testing/benchmarks/coverage/style/quality
        \end{itemize}
    \end{itemize}
  \item What is this paper:
    \begin{itemize}
      \item 4 codes, 2 FD 2 FE
      \item 2 models, simple one after MT paper, complex Marlim one
    \end{itemize}
  \item Outline of paper
  \item Zenodo, provide all codes to reproduce the results.
\end{itemize}

~\\ \hrule

There are many codes for EM - our focus here is on the application to
controlled-source electromagnetic measurements, even though all of the codes
can handle different setups too.

in such fields as the modelling of communication systems (antennas, radar,
satellites), medical imaging, and others.

1D, 2D, we look at 3D.
time domain and frequency domain

Common methods in geophysics are the

\begin{itemize}
  \item integral equation (IE, \cite{GJI.74.Raiche, GEO.06.Zhdanov,
    CAG.16.Kruglyakov}), (scattering equation) (anomalies in layered
    background), computational effort comparably low
  \item differential equation such as finite difference (FD \cite{IEEE.66.Yee,
    GEO.93.Wang}), simple to implement
  \item and finite volume (FV \cite{EM.90.Madsen, GEO.14.Jahandari}),
  \item or and finite element (FE \cite{GEO.04.Commer GEO.09.Streich})
    (approximating the DE on edges and nodals), good for complex geometries,
  \item finite integration technique (FIT, \cite{AEU.77.Weiland},
    \cite{PIER.01.Clemens}, \cite{GP.06.Mulder})
\end{itemize}

OctTree meshes or severe stretching \citep{ECP.07.Haber}

Elmer FEM multiphysical simulation software P. Råback, P.-L. Forsström, M. Lyly
and M. Gröhn, Elmer - finite element package for the solution of partial
differential equations, poster presentation, EGEE User Forum, 2007, Manchester,
UK. http://www.csc.fi/elmer

Many commercial, e.g., COMSOL Multiphysics; EMGS, RSI, KMS; the oil and service
companies.

Consortia (CEMI; Key)

Universities, e.g. Pottsdam (Streich, Grayver)

Computational cost IE < FD < FV < FE

Direct solvers vs indirect solvers, both its advantages and disadvantages;
memory, various sources.

Grids (regular grids, unstructured grids, Octree, tetrahedra, hexahedral,
adaptive, multigrids).

Boundary condition

@@

@@

The landscape in 3D CSEM modelling greatly changed in the last five years or
so. While before there were only closed-source codes owned by companies or
consortia (e.g., CEMI), or codes that you had to obtain from the authors, often
without much documentation, there was recently a wave of openly released codes.

@@

The landscape in 3D CSEM modelling greatly changed in the last five years or
so. While before there were only closed-source codes owned by companies or
consortia (e.g., CEMI) there was recently a wave of openly released codes.

@@

>>>>

Controlled-source electromagnetic (CSEM) measurements are a frequently applied
method in various geophysical exploration fields, such as geothermal and ground
water, oil and gas, mining, civil engineering, or geo-hazards. Modelling these
electromagnetic fields is therefore of great interest in order to understand
the measured data. Publications regarding 3D modelling in electromagnetic
methods started to appear as early as the 1970's and 1980's. These early
publications where integral equation (IE) methods, having an anomaly embedded
within a layered medium, mostly for loop-loop type transient EM measurements
\citep{GJI.74.Raiche, GEO.75.Hohmann, GJI.82.Das, GEO.86.Newman} and
magnetotelluric (MT) measurements \citep{GEO.84.Wannamaker}.

In the 1990's computer became sufficiently powerful that 3D modelling gained
traction, which resulted amongst other in the publication of the book
\emph{Three-Dimensional Electromagnetics} by the SEG
\citep{B.SEG.99.Oristaglio}. Often cited publications from that time are
\cite{RSC.94.Mackie}, 3D MT computation; \cite{RS.94.Druskin}, frequency- and
time-domain modelling using a Yee grid and a global Krylov subspace
approximation; and \cite{RS.96.Alumbaugh, GJI.97.Newman}, low-to-high frequency
computation on massively parallel computers.

The continuous improvement of computing power and the CSEM boom in the early
2000's in the hydrocarbon industry led to a wealth of publications. The amount
of available numerical solutions can be overwhelming and is part of the reason
why there are hundreds of publications about the topic. There are the different
methods to solve Maxwell's equation, such as the IE method
\citep{GJI.74.Raiche, RS.02.Hursan, GEO.06.Zhdanov, GP.10.Tehrani,
CAG.16.Kruglyakov, MGS.17.Kruglyakov} and different variations of the
differential equation (DE) method, for instance finite differences (FD)
\citep{IEEE.66.Yee, GEO.93.Wang, RSC.94.Mackie, RS.94.Druskin, GEO.09.Streich,
CAG.13.Sommer}, finite elementes (FE) \citep{GJI.11.Schwarzbach, GEO.04.Commer,
GEO.12.daSilva, GJI.13.Puzyrev, GJI.13.Grayver, SEG.16.Zhang}, finite volume
(FV) \citep{EM.90.Madsen, ECP.07.Haber, GEO.14.Jahandari}, and finite
integration technique (FIT) \citep{PIER.01.Clemens, GP.06.Mulder}. And these
are just the most common ones.

@@

A very well written overview up to the year 2005 of the different approaches to
3D EM modelling is given by \cite{SG.05.Avdeev}. In the last 15 years the
publications with regards to 3D EM  modelling grew tremendously, driven
by the ever increasing computing power. Avdeev finished his review with the
following statement: «\emph{The most important challenge that faces the EM
community today is to convince software developers to put their 3-D EM forward
and inverse solutions into the public domain, at least after some time. This
would have a strong impact on the whole subject and the developers would
benefit from feedback regarding the real needs of the end-users.}»

It is this topic that we address, open-source 3D CSEM codes. Some major
exploration service companies as well as certain research consortia and
research groups have their 3D CSEM codes. But how about codes that are
open-source? Traditionally research code was often available upon request by
the author. Codes distributed that way often come with the request to not share
the code, or they often come without a license attached at all (which means
that they are not open-source). Also, at times there can be a significant
hurdle to install a code. However, the meaning of open-source evolved rapidly
in the last decades. Today, open-source generally not only means that a code is
available (with a proper license). It includes much more today, such as that
the code is hosted online, is under version control, has the possibility to
file issues and make pull requests to fix bugs. Well maintained codes often
have continuous integration that includes testing of the code base and online
hosted documentation. As such the term has shifted from purely open-source code
to development in the open and increasingly building a global community. There
are a number of projects within the realm of geophysical exploration, besides
the ones we present here, \todo{Others coming to your mind?}e.g. pyGIMLi
\citep{CAG.17.Rucker} and Fatiando \citep{JOSS.18.Uieda}. Newer \todo{wrong
term, think about it}programming environments  have also brought much
simplifications in the installation process and have most importantly
simplified it for code developers to make their codes easily available and
installable. All our presented codes are in the Python ecosystem, and each code
can be installed with a single command (which includes downloading and
installation).

A second topic we want to address is verification. New codes are generally
verified against analytical solutions. However, analytical and semi-analytical
solutions only exist for very simple models. There is no possibility to check
how good a computed response is, expect for comparing the result from different
modellers. If different modellers using different discretisation and
implementations of Maxwell's equations yield the same result it can give
confidence in the accuracy.

It is worth mentioning that there are many more 3D EM codes developed in other
fields than geophysics such as communication systems (antennas, radar,
satellites), medical imaging. There are various open-source codes in these
fields \todo{Add at least three references} too, e.g., Elmer
(\href{http://www.csc.fi/elmer}{csc.fi/elmer}, couldn't find a publication,
just a poster). While these codes could potentially be used for the same goals
as presented here we restrict our review to codes purpose-built for geophysical
applications.

In the following section we introduce the codes under consideration. The
numerical result consider two cases. First a layered background with vertical
transverse isotropy containing three blocks, followed by a realistic, complex
model.

@@

Custom meshes for these lines.

@@

The landscape in 3D CSEM modelling greatly changed in the last five years or
so. While before there were only closed-source codes owned by companies or
consortia, or codes that you had to obtain from the authors, often
without much documentation, there was recently a wave of openly released codes.

Points to discuss and mention
\begin{itemize}
  \item Direct solvers vs indirect solvers, both its advantages and
    disadvantages; memory, various sources.
\end{itemize}

@@

The landscape in 3D CSEM modelling greatly changed in the last five years or
so. While before there were only closed-source codes owned by companies or
consortia there was recently a wave of openly released codes.

@@

\section{Data}

All files to reproduce the shown results and more are available at
\todo{zenodo}\dots.
